---
title: "R Notebook"
output: html_notebook
---


```{r}
pkgs <- sort(c('tidyverse', 'arules', 'arulesViz', 'mlbench'))

pkgs_install <- pkgs[!(pkgs %in% installed.packages()[,"Package"])]
if(length(pkgs_install)) install.packages(pkgs_install)
```

```{r}
library(tidyverse)
library(arules)
library(arulesViz)
```

```{r}
data(Zoo, package = "mlbench")
head(Zoo)
```
The data in the data.frame need to be converted into a set of transactions where each row represents a transaction and each column is translated into items. This is done using the constructor transactions(). For the Zoo data set this means that we consider animals as transactions and the different traits (features) will become items that each animal has. For example the animal antelope has the item hair in its transaction.
```{r}
trans <- transactions(Zoo)
```

The conversion gives a warning because only discrete features (factor and logical) can be directly translated into items. Continuous features need to be discretized first.

What is column 13?

```{r}
summary(Zoo[13])
```

```{r}
ggplot(Zoo, aes(legs)) + geom_bar()
```

```{r}
Zoo$legs |> table()
```

Possible solution: Make legs into has/does not have legs

```{r}
Zoo_has_legs <- Zoo |> mutate(legs = legs > 0)
ggplot(Zoo_has_legs, aes(legs)) + geom_bar()
```
Zoo_has_legs$legs |> table()

use each unique value as an item:

```{r}
Zoo_unique_leg_values <- Zoo |> mutate(legs = factor(legs))
Zoo_unique_leg_values$legs |> head()
```

```{r}
Zoo_discretized_legs <- Zoo |> mutate(
  legs = discretize(legs, breaks = 2, method="interval")
)
table(Zoo_discretized_legs$legs)
```

Convert data into a set of transactions

```{r}
trans <- transactions(Zoo_has_legs)
trans
```

```{r}
summary(trans)
```

```{r}
colnames(trans)
```

```{r}
as(trans, "matrix")[1:3,]
```
Look at the transactions as sets of items

```{r}
inspect(trans[1:3])
```
```{r}
image(trans)
```
Look at the relative frequency (=support) of items in the data set. Here we look at the 10 most frequent items.

```{r}
itemFrequencyPlot(trans,topN = 20)
```
```{r}
ggplot(
  tibble(
    Support = sort(itemFrequency(trans, type = "absolute"), decreasing = TRUE),
    Item = seq_len(ncol(trans))
  ), aes(x = Item, y = Support)) + geom_line()
```

```{r}
Zoo_factors <- Zoo_has_legs |> mutate(across(where(is.logical), factor))
sapply(Zoo_factors, class)
```

```{r}
summary(Zoo_factors)
```

```{r}
trans_factors <- transactions(Zoo_factors)
trans_factors
```

```{r}
itemFrequencyPlot(trans_factors, topN = 20)
```
```{r}
## Select transactions that contain a certain item
trans_insects <- trans_factors[trans %in% "type=insect"]
trans_insects
```

```{r}
inspect(trans_insects)
```
Vertical Layout (Transaction ID Lists)
The default layout for transactions is horizontal layout (i.e. each transaction is a row). The vertical layout represents transaction data as a list of transaction IDs for each item (= transaction ID lists).

```{r}
vertical <- as(trans, "tidLists")
as(vertical, "matrix")[1:10, 1:5]
```
Frequent Itemsets
Mine Frequent Itemsets
For this dataset we have already a huge number of possible itemsets

```{r}
2^ncol(trans)
```

Find frequent itemsets (target=“frequent”) with the default settings.

```{r}
its <- apriori(trans, parameter=list(target = "frequent"))
```
Default minimum support is .1 (10%). Note: We use here a very small data set. For larger datasets the default minimum support might be to low and you may run out of memory. You probably want to start out with a higher minimum support like .5 (50%) and then work your way down.

```{r}
its
```

```{r}
5/nrow(trans)
```

In order to find itemsets that effect 5 animals I need to go down to a support of about 5%.

```{r}
its <- apriori(trans, parameter=list(target = "frequent", support = 0.05))
```
```{r}
its
```

```{r}
its <- sort(its, by = "support")
its |> head(n = 10) |> inspect()
```
```{r}
ggplot(tibble(`Itemset Size` = factor(size(its))), aes(`Itemset Size`)) + 
  geom_bar()
```
```{r}
its[size(its) > 8] |> inspect()
```
Concise Representation of Itemsets
Find maximal frequent itemsets (no superset if frequent)

```{r}
its_max <- its[is.maximal(its)]
its_max
```

```{r}
its_max |> head(by = "support") |> inspect()
```
Find closed frequent itemsets (no superset if frequent)

```{r}
its_closed <- its[is.closed(its)]
its_closed
```

```{r}
its_closed |> head(by = "support") |> inspect()
```

```{r}
counts <- c(
  frequent=length(its),
  closed=length(its_closed),
  maximal=length(its_max)
)

ggplot(as_tibble(counts, rownames = "Itemsets"),
  aes(Itemsets, counts)) + geom_bar(stat = "identity")
```
```{r}
rules <- apriori(trans, parameter = list(support = 0.05, confidence = 0.9))
```
Association Rules
Mine Association Rules
We use the APRIORI algorithm

```{r}
length(rules)
```

```{r}
rules |> head() |> inspect()
```

```{r}
rules |> head() |> quality()
```
Look at rules with highest lift

```{r}
rules <- sort(rules, by = "lift")
rules |> head(n = 10) |> inspect()
```
Create rules using the alternative encoding (with “FALSE” item)

```{r}
r <- apriori(trans_factors)
```
```{r}
r
```

```{r}
print(object.size(r), unit = "Mb")
```

```{r}
inspect(r[1:10])
```
```{r}
r |> head(n = 10, by = "lift") |> inspect()
```
```{r}
interestMeasure(rules[1:10], measure = c("phi", "gini"),
  trans = trans)
```
```{r}
quality(rules) <- cbind(quality(rules),
  interestMeasure(rules, measure = c("phi", "gini"),
    trans = trans))
```

Find rules which score high for Phi correlation

```{r}
rules |> head(by = "phi") |> inspect()
```
Mine Using Templates
Sometimes it is beneficial to specify what items should be where in the rule. For apriori we can use the parameter appearance to specify this. In the following we restrict rules to an animal type in the RHS and any item in the LHS.


```{r}
type <- grep("type=", itemLabels(trans), value = TRUE)
type
```

```{r}
rules_type <- apriori(trans, appearance= list(rhs = type))
```
```{r}
rules_type |> sort(by = "lift") |> head() |> inspect()
```
```{r}
plot(rules)
```

```{r}
plot(rules, control = list(jitter = 0))
```

```{r}
plot(rules, shading = "order")
```
```{r}
plot(rules, method = "grouped")
```

```{r}
plot(rules, method = "graph")
```
```{r}
plot(rules |> head(by = "phi", n = 100), method = "graph")
```
```{r}
plot(rules, method = "matrix", engine = "html") 
```

