---
title: "R Notebook"
output: html_notebook
---
```{r}
pacman::p_load('tidyverse', 'tidymodels', 'skimr', 'here', 'plotly', 'factoextra', 'cluster')

```

```{r}
library(tidyverse)

# Read the csv file into a tibble
seeds <- read_csv(file = "https://raw.githubusercontent.com/MicrosoftDocs/ml-basics/master/data/seeds.csv")

# Print the first 10 rows of the data
seeds %>% 
  slice_head(n = 5)

```

```{r}
# Explore dimension and type of columns
seeds %>% 
  glimpse()

```

```{r}
library(skimr)

# Obtain Summary statistics
seeds %>% 
  skim()
D
```

```{r}
# Narrow down to desired features
seeds_select <- seeds %>% 
  select(!groove_length) %>% 
  mutate(species = factor(species))

# View first 5 rows of the data
seeds_select %>% 
  slice_head(n = 5)

```
As you can see, we now have six data points (or features) for each instance (observation) of a seed’s species. So you could interpret these as coordinates that describe each seed’s location in six-dimensional space.

Now, of course six-dimensional space is difficult to visualise in a three-dimensional world, or on a two-dimensional plot; so we’ll take advantage of a mathematical technique called Principal Component Analysis (PCA) to analyze the relationships between the features and summarize each observation as coordinates for two principal components - in other words, we’ll translate the six-dimensional feature values into two-dimensional coordinates.

Principal Component Analysis (PCA) is a dimension reduction method that aims at reducing the feature space, such that, most of the information or variability in the data set can be explained using fewer uncorrelated features.

PCA works by receiving as input P variables (in this case six) and calculating the normalized linear combination of the P variables. This new variable is the linear combination of the six variables that captures the greatest variance out of all of them. PCA continues to calculate other normalized linear combinations but with the constraint that they need to be completely uncorrelated to all the other normalized linear combinations.
```{r}
#PCA works well when the variables are normalized (centered and scaled)
# Specify a recipe for pca
pca_rec <- recipe(~ ., data = seeds_select) %>% 
  update_role(species, new_role = "ID") %>% 
  step_normalize(all_predictors()) %>% 
  step_pca(all_predictors(), num_comp = 2, id = "pca")

# Print out recipe
pca_rec


```

Compared to supervised learning techniques, we have no outcome variable in this recipe.

By updating the role of the species column to ID, this tells the recipe to keep the variable but not use it as either an outcome or predictor.

By calling prep() which estimates the statistics required by PCA and applying them to seeds_features using bake(new_data = NULL), we can get the fitted PC transformation of our features.



```{r}
# Estimate required statistcs 
pca_estimates <- prep(pca_rec)

# Return preprocessed data using bake
features_2d <- pca_estimates %>% 
  bake(new_data = NULL)

# Print baked data set
features_2d %>% 
  slice_head(n = 5)

```

These two components capture the maximum amount of information (i.e. variance) in the original variables. From the output of our prepped recipe pca_estimates, we can examine how much variance each component accounts for:



```{r}
# Examine how much variance each PC accounts for
pca_estimates %>% 
  tidy(id = "pca", type = "variance") %>% 
  filter(str_detect(terms, "percent"))

```

```{r}
theme_set(theme_light())
# Plot how much variance each PC accounts for
pca_estimates %>% 
  tidy(id = "pca", type = "variance") %>% 
  filter(terms == "percent variance") %>% 
  ggplot(mapping = aes(x = component, y = value)) +
  geom_col(fill = "midnightblue", alpha = 0.7) +
  ylab("% of total variance")

```

This output tibbles and plots shows how well each principal component is explaining the original six variables. For example, the first principal component (PC1) explains about 72% of the variance of the six variables. The second principal component explains an additional 16.97%, giving a cumulative percent variance of 89.11%. This is certainly better. It means that the first two variables seem to have some power in summarizing the original six variables.

Naturally, the first PC (PC1) captures the most variance followed by PC2, then PC3, etc.

Now that we have the data points translated to two dimensions PC1 and PC2, we can visualize them in a plot:



```{r}
# Visualize PC scores
features_2d %>% 
  ggplot(mapping = aes(x = PC1, y = PC2)) +
  geom_point(size = 2, color = "dodgerblue3")

```
Hopefully you can see at least two, arguably three, reasonably distinct groups of data points; but here lies one of the fundamental problems with clustering - without known class labels, how do you know how many clusters to separate your data into?

One way we can try to find out is to use a data sample to create a series of clustering models with an incrementing number of clusters, and measure how tightly the data points are grouped within each cluster. A metric often used to measure this tightness is the within cluster sum of squares (WCSS), with lower values meaning that the data points are closer. You can then plot the WCSS for each model.

We’ll use the built-in kmeans() function, which accepts a data frame with all numeric columns as it’s primary argument to perform clustering - means we’ll have to drop the species column. For clustering, it is recommended that the data have the same scale. We can use the recipes package to perform these transformations.

```{r}
# Drop target column and normalize data
seeds_features<- recipe(~ ., data = seeds_select) %>% 
  step_rm(species) %>% 
  step_normalize(all_predictors()) %>% 
  prep() %>% 
  bake(new_data = NULL)

# Print out data
seeds_features %>% 
  slice_head(n = 5)

```
Now, let’s explore the WCSS of different numbers of clusters.

We’ll get to use map() from the purrr package to apply functions to each element in list.

map() functions allow you to replace many for loops with code that is both more succinct and easier to read. The best place to learn about the map() functions is the iteration chapter in R for data science.

broom::augment.kmeans() accepts a model object and returns a tibble with exactly one row of model summaries. The summaries are typically goodness of fit measures, p-values for hypothesis tests on residuals, or model convergence information.

```{r}
set.seed(2056)
# Create 10 models with 1 to 10 clusters
kclusts <- tibble(k = 1:10) %>% 
  mutate(
    model = map(k, ~ kmeans(x = seeds_features, centers = .x, nstart = 20)),
    glanced = map(model, glance)) %>% 
  unnest(cols = c(glanced))

# View results
kclusts

```

```{r}
# Plot Total within-cluster sum of squares (tot.withinss)
kclusts %>% 
  ggplot(mapping = aes(x = k, y = tot.withinss)) +
  geom_line(size = 1.2, alpha = 0.5, color = "dodgerblue3") +
  geom_point(size = 2, color = "dodgerblue3")

```

We seek to minimize the the total within-cluster sum of squares, by performing K-means clustering. The plot shows a large reduction in WCSS (so greater tightness) as the number of clusters increases from one to two, and a further noticable reduction from two to three clusters. After that, the reduction is less pronounced, resulting in an elbow 💪in the chart at around three clusters. This is a good indication that there are two to three reasonably well separated clusters of data points.

The algorithm we used to approximate the number of clusters in our data set is called K-Means. Let’s get to the finer details, shall we?

K-Means is a commonly used clustering algorithm that separates a dataset into K clusters of equal variance such that observations within the same cluster are as similar as possible (i.e., high intra-class similarity), whereas observations from different clusters are as dissimilar as possible (i.e., low inter-class similarity). The number of clusters, K, is user defined.

The basic algorithm has the following steps:

Specify the number of clusters to be created (this is done by the analyst). Taking the flowers example we used at the beginning of the lesson, this means deciding how many clusters you want to use to group the flowers.
Next, the algorithm randomly selects K observations from the data set to serve as the initial centers for the clusters (i.e., centroids).
Next, each of the remaining observations (in this case flowers) are assigned to its closest centroid.
Next, the new means of each cluster is computed and the centroid is moved to the mean.
Now that the centers have been recalculated, every observation is checked again to see if it might be closer to a different cluster. All the objects are reassigned again using the updated cluster means. The cluster assignment and centroid update steps are iteratively repeated until the cluster assignments stop changing (i.e., when convergence is achieved). Typically, the algorithm terminates when each new iteration results in negligible movement of centroids and the clusters become static.
Note that due to randomization of the initial k observations used as the starting centroids, we can get slightly different results each time we apply the procedure. For this reason, most algorithms use several random starts and choose the iteration with the lowest WCSS. As such, it is strongly recommended to always run K-Means with several values of nstart to avoid an undesirable local optimum.
So training usually involves multiple iterations, reinitializing the centroids each time, and the model with the best (lowest) WCSS is selected. The following animation shows this process:

The algorithm we used to approximate the number of clusters in our data set is called K-Means. Let’s get to the finer details, shall we?

K-Means is a commonly used clustering algorithm that separates a dataset into K clusters of equal variance such that observations within the same cluster are as similar as possible (i.e., high intra-class similarity), whereas observations from different clusters are as dissimilar as possible (i.e., low inter-class similarity). The number of clusters, K, is user defined.

The basic algorithm has the following steps:

Specify the number of clusters to be created (this is done by the analyst). Taking the flowers example we used at the beginning of the lesson, this means deciding how many clusters you want to use to group the flowers.
Next, the algorithm randomly selects K observations from the data set to serve as the initial centers for the clusters (i.e., centroids).
Next, each of the remaining observations (in this case flowers) are assigned to its closest centroid.
Next, the new means of each cluster is computed and the centroid is moved to the mean.
Now that the centers have been recalculated, every observation is checked again to see if it might be closer to a different cluster. All the objects are reassigned again using the updated cluster means. The cluster assignment and centroid update steps are iteratively repeated until the cluster assignments stop changing (i.e., when convergence is achieved). Typically, the algorithm terminates when each new iteration results in negligible movement of centroids and the clusters become static.
Note that due to randomization of the initial k observations used as the starting centroids, we can get slightly different results each time we apply the procedure. For this reason, most algorithms use several random starts and choose the iteration with the lowest WCSS. As such, it is strongly recommended to always run K-Means with several values of nstart to avoid an undesirable local optimum.
So training usually involves multiple iterations, reinitializing the centroids each time, and the model with the best (lowest) WCSS is selected. 

Now, back to our seeds example. After creating a series of clustering models with different numbers of clusters and plotting the WCSS across the clusters, we noticed a bend at around k = 3. This bend indicates that additional clusters beyond the third have little value and that there are two to three reasonably well separated clusters of data points.

So, let’s perform K-Means clustering specifying k = 3 clusters and add the classifications to the data set using augment.



```{r}
set.seed(2056)
# Fit and predict clusters with k = 3
final_kmeans <- kmeans(seeds_features, centers = 3, nstart = 100, iter.max = 1000)

# Add cluster prediction to the data set
results <- augment(final_kmeans, seeds_features) %>% 
# Bind pca_data - features_2d
  bind_cols(features_2d)

results %>% 
  slice_head(n = 5)

```

Let’s see those cluster assignments with the two dimensional data points. We’ll add some touch of interactivity using the plotly package, so feel free to hover.

```{r}
# Plot km_cluster assignmnet on the PC data
cluster_plot <- results %>% 
  ggplot(mapping = aes(x = PC1, y = PC2)) +
  geom_point(aes(shape = .cluster), size = 2) +
  scale_color_manual(values = c("darkorange","purple","cyan4"))

# Make plot interactive
ggplotly(cluster_plot)

```
🤩🤩 Hopefully, the data has been separated into three distinct clusters.

So what’s the practical use of clustering? In some cases, you may have data that you need to group into distict clusters without knowing how many clusters there are or what they indicate. For example a marketing organization might want to separate customers into distinct segments, and then investigate how those segments exhibit different purchasing behaviors.

Sometimes, clustering is used as an initial step towards creating a classification model. You start by identifying distinct groups of data points, and then assign class labels to those clusters. You can then use this labelled data to train a classification model.

In the case of the seeds data, the different species of seed are already known and encoded as 0 (Kama), 1 (Rosa), or 2 (Canadian), so we can use these identifiers to compare the species classifications to the clusters identified by our unsupervised algorithm

```{r}
# Plot km_cluster assignmnet on the PC data
clust_spc_plot <- results %>% 
  ggplot(mapping = aes(x = PC1, y = PC2)) +
  geom_point(aes(shape = .cluster, color = species), size = 2, alpha = 0.8) +
  scale_color_manual(values = c("darkorange","purple","cyan4"))

# Make plot interactive
ggplotly(clust_spc_plot)

```
There may be some differences between the cluster assignments and class labels as shown by the different colors (species) within each cluster (shape). But the K-Means model should have done a reasonable job of clustering the observations so that seeds of the same species are generally in the same cluster. 💪


